%% acknowledgments.tex

% From mitthesis package
% Version: 1.02, 2024/06/19
% Documentation: https://ctan.org/pkg/mitthesis

\chapter*{Acknowledgments}
% \pdfbookmark[0]{Acknowledgments}{acknowledgments}
\addcontentsline{toc}{chapter}{Acknowledgements}

First, I would really like to thank my advisor, Professor Chuchu Fan: you’re very inspirational to me and there are many things that you’ve said that I’ve applied in my daily life which helped a lot. I’m also very grateful to Anjali Parashar, for all of the research and personal discussions, and for all the help along the way. I probably wouldn’t be writing this thesis if you both hadn’t brought me into this lab and supported me all the way through.

I’d also like to thank people from near the end of my time here: Professor Max Li, along with Micah Borrero and James Jones; and from near the beginning: Professor Mengjia Yan and Shixin Song; you all were great to work with and I really learned a lot. Charles Dawson: we’ve literally never met but I’m a huge fan so I’d like to thank you for existing. I’d like to express my immense gratitude to everyone in REALM as well: you all were really nice, and also let me have my own desk in the office – I’ve never had one before so it meant a lot.

Next, I’d like to thank the amazing undergraduate researchers that I’m lucky to have had the opportunity to mentor … is what I would say if I had any, but I don’t. Moving on, I am also grateful for Jaeyoun's Crazyflie experiments: after seeing those for a while I suddenly felt like baby-sitting the UR3e wasn’t so bad after all, but I hope they’ve been going well. Still, I would also like to take a moment to not express my gratitude for the UR3e. While we’re on the topic of inanimate objects: the low table I inherited changed my life – I don’t own a chair so it was nice to have something to work on the floor with. I am also grateful to my family for all of their support, as usual. I almost forgot: Neal, I didn’t know good class project partners existed but I guess I was wrong.

Lastly, I must express my heartfelt gratitude to Cindy, Ella, Jared, Rina, Tananya, and other friends here for everything: you’ve all made my time here far better than I thought it’d be and I can never thank you enough. Finally, I’d like to thank Ningning, my fish: words will never be enough to express how much I appreciate you for always being there for me (though I guess you do live in a fish tank and can’t really go anywhere).

